Longitudinal Principal Manifold Estimation

### 2024-04-05 paper/smi_2024.tex
add smi paper adaptation

\usepackage[
backend=biber,
style=authoryear,
url=false,
isbn=false,
date=year,
doi=false
]{biblatex}

@@

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{bm}

@@

\usepackage{csquotes}
\usepackage[english]{babel}

\renewcommand{\baselinestretch}{1.0}
\setlength{\oddsidemargin}{-0.5cm}
\setlength{\evensidemargin}{-0.5cm}
\renewcommand{\topmargin}{-2cm}
\renewcommand{\oddsidemargin}{0mm}
\renewcommand{\evensidemargin}{0mm}
\renewcommand{\textwidth}{180mm}
\renewcommand{\textheight}{240mm}

@@

Alzheimer's disease (AD) is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging (MRI) data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is a wide variability of shape and volume estimates of brain regions of interest over time in longitudinal studies resulting in major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This smoothing mitigates the effects of random disturbances to the manifold between time points. A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We show that the proposed method improves the estimation of surfaces of hippocampi and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}). Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}). For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with AD or cognitive impairment related to other causes.

Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. When analyzing data from studies with a large number of images, this approach may impose prohibitive costs. Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}). Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}). While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained. In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}). To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.

Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.

Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}). Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}; \cite{gao2024bi}). In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

@@

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection. In addition, applying the methods in high-dimensional settings is problematic. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $t \to \left( \frac{\partial ^2 f}{\partial t_i \partial t_j}(t) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, denoted $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows.

@@

where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, the set $\mathcal{D}'(\mathbb{R}^d)$ denotes the collection of generalized functions defined on $\mathbb{R}^d$ (also known as ``distributions" in the mathematical community, see Chapter 6 of \cite{rudin1991functional} for details), and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.

@@

The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. In this section, we introduce an approach to extending this framework to longitudinal clouds of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. We consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e. $D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest. Now consider that the data are observed longitudinally during several visits by a study participant. Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit. Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$. The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.4]{sim_case1_plot2}
  \caption{Longitudinal Simulated Data with $D = 2$. Data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are simulated at 5 time points. Random noise is added both in time and space as $\zeta \mathbf{g}(t) + \iota$, where $\mathbf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, and $\iota$ represents normally distributed noise applied to each point individually.}
  \label{fig:sample_data}
\end{figure}

Assume that our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth. It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, i.e. modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc. The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method above then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time. In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$. In order to address the limitations of the existing approaches for the setting described above, we now propose longitudinal principal manifold estimation (LPME) algorithm to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.

@@

\begin{figure}[ht]
  \centering
  \subfloat[\centering Data]{\label{fig:lpme_step1} \includegraphics[width=8cm]{sim_case1_plot2}}%
  \hfill
  \subfloat[\centering Step 1: Sample Size Reduction]{\label{fig:lpme_step2} \includegraphics[width=8cm]{lpme_data_reduction}}
  \vfill
  \subfloat[\centering Step 2: Initialization]{\label{fig:lpme_step3} \includegraphics[width=8cm]{lpme_initialization}}
  \hfill
  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=8cm]{lpme_fitting}}
  \caption{LPME Algorithm Steps. In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface. }
  \label{fig:lpme_steps}
\end{figure}

@@

The next step in the algorithm for estimation of the longitudinal principal manifold is initialization. Different algorithms use different approaches for this step. For example, the PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, implementation of commonly used approaches for initialization without considering the nuances posed by longitudinal data can be problematic. For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e. the brain region of interest at each time point) is likely to result in inconsistent parameterization directions across time points (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another). This is demonstrated in Figure \ref{fig:inconsistent_parameterization}, which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:sample_data}. Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.

\begin{figure}[h]
  \centering
  \includegraphics[height=7cm]{inconsistent_parameterization}
  \caption{Simulated data with $D = 2$. The curves shown in this plot represents embedding maps estimated by PME for each time point, with time indicated by the color of the curve. The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates. Randomness in the initialization of the PME algorithm may result in misaligned parameterizations of the estimated embedding maps. For most functions, the $X$ values increase with increasing values for $R$. However, two functions have $X$ values that decrease with increasing $R$. The coefficients for these two functions deviate enough from the other function coefficients that smoothing over the functions with respect to time becomes much more difficult.}
  \label{fig:inconsistent_parameterization}
\end{figure}

@@

Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds. Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency condition introduced in \cite{hastiePrincipalCurves1989}. To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.

\begin{figure}[ht]
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
  \caption{Illustration of our proposed data augmentation approach using the Unit Circle. The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
  \label{fig:unit_circle_augmentation}
\end{figure}

@@

Visualizations of the results from some of the simulation cases, truncated for concision, are shown in Figures \ref{fig:sim_case1} and \ref{fig:sim_case6}. Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time. Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.

@@

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:lhipp_cross_sections}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.

>>>>

\usepackage[
backend=biber,
style=ext-authoryear,
url=false,
isbn=false,
date=year,
doi=false,
eprint=false,
articlein=false
]{biblatex}

%\renewbibmacro{in:}{}
%\DeclareFieldFormat{pages}{#1}

@@

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{bm}
\usepackage[margin=0.7in]{geometry}
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{0.1\baselineskip}{0.2\baselineskip}

@@

\usepackage{csquotes}
\usepackage[english]{babel}
%
% \renewcommand{\baselinestretch}{1.0}
% \setlength{\oddsidemargin}{-0.5cm}
% \setlength{\evensidemargin}{-0.5cm}
% \renewcommand{\topmargin}{-2cm}
% \renewcommand{\oddsidemargin}{0mm}
% \renewcommand{\evensidemargin}{0mm}
% \renewcommand{\textwidth}{180mm}
% \renewcommand{\textheight}{240mm}

@@

Alzheimer's disease (AD) is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging (MRI) data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is a wide variability of shape and volume estimates of brain regions of interest over time in longitudinal studies resulting in major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This smoothing mitigates the effects of random disturbances to the manifold between time points. A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We apply the proposed method to estimate the volumes of hippocampi and thalamuses of participants in the Alzheimer's Disease Neuroimaging Initiative dataset.

@@

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}). Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}). For example, it is common to observe atrophy in the hippocampus of those with AD or cognitive impairment related to other causes.

Image segmentation is the process by which subcortical structures are extracted from MRI images of the brain, allowing the identification of differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation. However, this approach is highly time and resource intensive, and may impose prohibitive costs when analyzing large numbers of images. Additionally, manual segmentation suffers from meaningful inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}). Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}). While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

Failures of automatic segmentation algorithms may have substantial effects when analyzing large datasets of images where manual quality control may not be feasible. In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}).

Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach used. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.

Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}). In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data and include it in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach that estimates manifolds over time while maintaining the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then impose smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

@@

The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, this approach has several limitations, and applying the methods in high-dimensional settings is problematic (\cite{duchamp1996extremal}). Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework.

In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $t \to \left( \frac{\partial ^2 f}{\partial t_i \partial t_j}(t) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, denoted $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows.

@@

where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, the set $\mathcal{D}'(\mathbb{R}^d)$ denotes the collection of generalized functions defined on $\mathbb{R}^d$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.

@@

The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. In this section, we introduce an approach to extending this framework to longitudinal clouds of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. We consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e. $D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest. Now consider that the data are observed longitudinally during several visits by a study participant. Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit. Figure \ref{fig:sim_case1} displays a simplified example of this setting using simulated data where $D = 2$. The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.

% \begin{figure}[h]
%   \centering
%   \includegraphics[scale=0.4]{sim_case1_plot2}
%   \caption{Longitudinal Simulated Data with $D = 2$. Data from the functions $x_1 = r$ and $x_2 = \alpha\sin(\beta r + \frac{\pi}{2})$ are simulated at 5 time points. Random noise is added both in time and space as $\zeta \mathbf{g}(t) + \iota$, where $\mathbf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time point, and $\iota$ represents normally distributed noise applied to each point individually.}
%   \label{fig:sample_data}
% \end{figure}
%
Assume that our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth. It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, i.e. modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc. The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method above then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time. In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$. In order to address the limitations of the existing approaches for the setting described above, we now propose longitudinal principal manifold estimation (LPME) algorithm to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.

@@

\begin{figure}[ht]
  \centering
  \subfloat[\centering Data]{\label{fig:sim_case1} \includegraphics[width=8cm]{sim_case1_plot2}}%
  \hfill
  \subfloat[\centering Step 1: Sample Size Reduction]{\label{fig:lpme_step2} \includegraphics[width=8cm]{lpme_data_reduction}}
  \vfill
  \subfloat[\centering Step 2: Initialization]{\label{fig:lpme_step3} \includegraphics[width=8cm]{lpme_initialization}}
  \hfill
  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=8cm]{lpme_fitting}}
  \caption{LPME Algorithm Steps. In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface. }
  \label{fig:lpme_steps}
\end{figure}

@@

The next step in the algorithm for estimation of the longitudinal principal manifold is initialization. Different algorithms use different approaches for this step. For example, the PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, implementation of commonly used approaches for initialization without considering the nuances posed by longitudinal data can be problematic. For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e. the brain region of interest at each time point) is likely to result in inconsistent parameterization directions across time points (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another). Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.

% \begin{figure}[h]
%   \centering
%   \includegraphics[height=7cm]{inconsistent_parameterization}
%   \caption{Simulated data with $D = 2$. The curves shown in this plot represents embedding maps estimated by PME for each time point, with time indicated by the color of the curve. The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates. Randomness in the initialization of the PME algorithm may result in misaligned parameterizations of the estimated embedding maps. For most functions, the $X$ values increase with increasing values for $R$. However, two functions have $X$ values that decrease with increasing $R$. The coefficients for these two functions deviate enough from the other function coefficients that smoothing over the functions with respect to time becomes much more difficult.}
%   \label{fig:inconsistent_parameterization}
% \end{figure}

@@

Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds. Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency condition introduced in \cite{hastiePrincipalCurves1989}. To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.

% \begin{figure}[ht]
%   \centering
%   \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=8.75cm]{unit_circle_D2}}}%
%   \hfill
%   \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=8.75cm]{unit_circle_D3}}}
%   \caption{Illustration of our proposed data augmentation approach using the Unit Circle. The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
%   \label{fig:unit_circle_augmentation}
% \end{figure}

@@

Visualizations of the results from some of the simulation cases, truncated for concision, are shown in Figure \ref{fig:sim_case1}. Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time. Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.

@@

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:lhipp_cross_sections}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:lthal_cross_sections}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.
